---
title: "Ricker and Larkin Models - Assignment 4"
output: github_document
author: "Christian Carson"
date: "03/05/2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Read in the data and set up the procedure section. 
Data can be read from github at: https://github.com/christiancarson/REM661/blob/main/Assingment%204/Sockeye.data

```{r}
data <- read.table("https://raw.githubusercontent.com/christiancarson/REM661/main/Assingment%204/Sockeye.data", header=TRUE, sep="\t")
stocks    <- unique( data$Stock )
nstock    <- length( stocks )
```

#\ln\left(\frac{R_t}{S_t}\right)=\alpha-\beta S_t+\omega_t

The Ricker model is written as:
$$\ln\left(\frac{R_t}{S_t}\right)=\alpha-\beta S_t+\omega_t$$

where:
- $R_t$ is the number of recruits in year $t$
- $S_t$ is the number of spawners in year $t$
- $\alpha$ is the intercept
- $\beta$ is the slope
- $\omega_t$ is the error term

Here we will fit a linear model to estimate alpha and beta in log( R/S ) = alpha - beta * S
```{r Ricker}
Ricker <- function(){
  
  # Create a list to fill with outputs from each stock
  rick <- list()
  results <- data.frame(Stock=character(), Alpha=numeric(), Beta=numeric(), stringsAsFactors=FALSE)
  
  # Loop over stocks
  for(i in 1:nstock){
    stock.id <- data$Stock == stocks[i]
    R <- data$Recruits[stock.id]
    S <- data$Spawners[stock.id]
    
    fit <- lm(log(R/S) ~ S)  # Linear regression on transformed Ricker model
    rick[[i]] <- list(fit=fit, r=exp(fit$fitted.values) * S)
    
    results <- rbind(results, data.frame(Stock=stocks[i], Alpha=coef(fit)[1], Beta=coef(fit)[2]))
  }
  
  return(list(models=rick, coefficients=results))
}
```
Next, we will create a function to estimate parameters and fitted values based on the Larkin model. The Larkin model is written as:
$$\ln\left(\frac{R_t}{S_t}\right)=\alpha-\beta_0S_t-\beta_1S_{t-1}-\beta_2S_{t-2}-\beta_3S_{t-3}+\omega_t$$

where:
- $R_t$ is the number of recruits in year $t$
- $S_t$ is the number of spawners in year $t$
- $\alpha$ is the intercept
- $\beta_0$ is the slope for the current year
- $\beta_1$ is the slope for the previous year
- $\beta_2$ is the slope for the year before the previous year
- $\beta_3$ is the slope for the year before the year before the previous year
- $\omega_t$ is the error term

```{r}
Larkin <- function(){
  # Create a list to fill with outputs from each stock
  lark <- list()
  results <- data.frame(Stock=character(), Alpha=numeric(), Beta0=numeric(), Beta1=numeric(), Beta2=numeric(), Beta3=numeric(), stringsAsFactors=FALSE)
  
  # Loop over stocks
  for(i in 1:nstock){
    stock.id <- data$Stock == stocks[i]
    R <- data$Recruits[stock.id]
    S <- data$Spawners[stock.id]
    nyr <- length(R)

    # Ensure enough data points for model fitting
    if(nyr > 3){
      Rt <- R[4:nyr]
      St <- S[4:nyr]
      Sm1 <- S[3:(nyr-1)]
      Sm2 <- S[2:(nyr-2)]
      Sm3 <- S[1:(nyr-3)]

      dat <- data.frame(Rt=Rt, St=St, Sm1=Sm1, Sm2=Sm2, Sm3=Sm3)
      fit <- glm(log(Rt/St) ~ St + Sm1 + Sm2 + Sm3, data=dat)

      lark[[i]] <- list(fit=fit, r=exp(fit$fitted.values) * St)
      
      coefs <- coef(fit)
      results <- rbind(results, data.frame(Stock=stocks[i], Alpha=coefs[1], Beta0=coefs[2], Beta1=coefs[3], Beta2=coefs[4], Beta3=coefs[5]))
    }
  }
  
  return(list(models=lark, coefficients=results))
}
```
fit1 <- Ricker()
fit2 <- Larkin()

