---
title: "Class 6 Nonlinear Estimation"
author: "Christian Carson"
date: "02/012/2024"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
````
$C_t = p N_0 - p K_t$
```{r}
#N_0 is the initial population or intercept
N_0 = 1000
#p is the proportion of the population that is captured or the slope
p = 0.1
#N_t is the population at time t
N_t <- vector()
#N_0 is the initial population
N_t[1] <- N_0
#C_t is the number of indviduals captured at time t
C_t <- vector()
K_t <- vector()
K_t[1] <- 0

for(t in 1:25) {
 C_t[t] <- p * N_t[t]
 K_t[t+1] <- K_t[t] + C_t[t]
    N_t[t+1] <- N_t[t]- C_t[t]
}

plot(C_t~K_t[1:25], ylim =c(0,N_0*p), xlim = c(0, N_0), yaxs = "i", type = "p", col = "blue", xlab = "Kt", ylab = "Ct")
lin_est <- lm(C_t~K_t[1:25])
abline(coef(lin_est), col = "red", lwd = 2, lty = 2)

p_hat <- -lin_est$coef[2]
N_0_hat <- lin_est$coef[1]/p_hat
#add p_hat and N_0_hat to the plot at middle right
text(0.5*N_0, 0.5*N_0*p, labels = paste("p_hat = ", round(p_hat, 2), "\n", "N_0_hat = ", round(N_0_hat, 2)), pos = 3)
````

```{r simulate 3-pass} 
"ThreePass" <- function(N_0 = 100, p=0.5) {
    C_t <- vector()
    N_t <- vector()
    K_t <- vector()
    K_t[1] <- 0
    N_t[1] <- N_0
    for(t in 1:3){
        C_t[t] <- rbinom(n=1, size = N_t[t],prob = p)
        K_t[t+1] <- K_t[t] + C_t[t]
        N_t[t+1] <- N_t[t] - C_t[t]
    }
    lin_est <- lm(C_t~K_t[1:3])
    N_0_hat <- lin_est$coef[1]/-lin_est$coef[2]

        return(N_0_hat)
}

```
This function generated catch values for a 3-pass removal model and estimates the initial population size. The function can be used to simulate data and estimate the initial population size. 

Now I want to run this repeatededly to evaluate bias and precision of the estimator. 

```{r repeated 3-pass}
N_0 <- 100
p <- 0.5
n_rep <- 100
N_0_vec <- vector()

for(i in 1:n_rep) {
    N_0_vec[i] <- ThreePass(N_0 = N_0, p = p)
}

#plot histogram of N_0_vec
hist(N_0_vec, main = "Histogram of N_0_vec", xlab = "N_0_hat", col = "black", freq = FALSE, breaks = 20)
```

## Schaeffer Production Model
The Schaeffer model is a nonlinear model that is used to estimate the intrinsic growth rate of a population. The model is given by the following equation:
$$
B_[t+1] = B_t + r B_t (1 - \frac{B_t}{K})
$$
where:

- $B_t$ is the biomass at time t

- r is the intrinsic growth rate

- K is the carrying capacity

Let's removes some animals from this population:

$$
B_[t+1] = B_t + r B_t (1 - \frac{B_t}{K})- H_t
$$

```{r simulate schaeffer}
"SimSchaeffer" <- function(Ht) {
    nT <- length(Ht)
    r <- 0.4
    K <- 1000000
    q <- 0.0001
    B0 <- K
    #process error
    v <- rnorm(n = nT, mean = 0, sd = 0.3)
    #observation error
    w <- rnorm(n = nT, mean = 0, sd = 0.1)
    #Biomass
    B_t <- vector()
    #
    It <- vector()
    #initial biomass
    B_t[1] <- B0
    for(t in 1:40) {
        It[t] <- B_t[t] * q * exp(w[t])
        B_t[t+1] <- B_t[t] + r * B_t[t] * (1 - B_t[t]/K) * exp(v[t]) - Ht[t] - It[t]
    }
    return(B_t)
}
```