---
title: "Assignment 5 - Prawns"
output: github_document
author: "Christian Carson"
date: "03/25/2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

This document models the population dynamics of spot prawns using Beverton-Holt recruitment model. The recruitment model equation is given by:

$$
N_{t1,t} = \frac{\alpha \cdot S_{pt-1}}{1 + \beta \cdot S_{pt-1}}
$$

where:
- \(N_{t1,t}\) is the number of new recruits at time \(t\),

- \(\alpha\) is the maximum recruitment rate,

- \(\beta\) is the rate at which recruitment increases with the number of spawning prawns,

- \(S_{pt-1}\) is the number of spawners at time \(t-1\).


```{r}
#total number of time steps, 3 months each covering 10 years. So (10*12)/3 = 40
nT <- 40
#create a function that will model the prawn population over time; MR is the mortality rate due to release after being caught, E is the fishing effort in terms of traps deployed
PrawnModel <- function(MR=0, E=400){
#base of the Beverton-holt model, where alpha is the maximum number of prawns that can be caught
alpha <- 1000
#beta is the rate at which the number of prawns caught increases with effort
beta <- 1e-3
#q is the catchability of the prawns
q <- 0.001
# assume 3-month time-steps
#p represents the transition rates between each life stage
p <- c( 1/5, 1/4, 1/3, 1, 1/2 )
#s represents the survival rates of each life stage, the last is 0 because the female prawns die after spawning
S <- c( 0.5, 0.55, 0.6, 0.6, 0.6, 0.0 )
#N0 represents the initial number of prawns in each life stage,the last is 0 because the female prawns die after spawning
N0 <- c( 1000000, 175000, 39000, 10000, 5000, 0 )
#seL represents the selectivity of the prawns at each life stage, or how likely they are to be caught
sel <- c( 0, 0.1, 0.3, 0.7, 1, 0.8 )
#prel represents the probability of a prawn being released after being caught, which modifies the survival rates below
prel <- c( 0, 0, 0, 0, 1, 0 )
#S represents the survival rates of each life stage. S is a product of 1 minus the catchabiliy multiplied by the effort, the selectivity, 1 minues the probability of being released, and plus the probability of mortality due to release
S <- S * (1- q * E * sel * ((1 - prel) + prel*MR ))
#setup the matrices to store the results for each time step
Nt <- matrix(nrow=nT,ncol=6)
Ct <- matrix(nrow=nT,ncol=6)
#initialize the first row of the matrices with the initial values from N0
Nt[1,] <- N0
#calculate the intial catch for the first time step based on the initial number of prawns, the catchability, the effort, and the selectivity
Ct[1,] <- Nt[1,]*q*E*sel
#setup the vectors for the spawner abundance and index, use "numeric" to ensure they dont get converted to logical
Spawn <- vector("numeric",nT)
SI <- vector("numeric",nT)
#initialize the spawning population and SI for the first time step, where spawn is Nt for the 5th stage, multiplied by the survival rate for the 5th stage, and the probability of being caught for the 5th stage
Spawn[1] <- Nt[1,5] * S[5] * p[5]
#initialize the spawner index for the first time step, which is the catch for the 5th stage divided by the effort (constant at 400)
SI[1] <- Ct[1,5] / E
#start the for loop for t in 2 to nT because we have already calculated the first time step, nT will go to 40 for the 3 month time steps
for (t in 2:nT) {
    # Beverton-Holt recruitment for new immatures based on spawners
    Nt[t,1] <- alpha * Spawn[t-1] / (1 + beta * Spawn[t-1])
    
    # Transitioning and surviving through life stages
    for (stage in 2:5) {
      Nt[t,stage] <- Nt[t-1,stage-1] * S[stage-1]
    }
    
    # Females after releasing eggs (typically don't survive long)
    Nt[t,6] <- Nt[t-1,5] * S[5]
    
    Ct[t,] <- Nt[t,] * q * E * sel  # Calculate catch
    Spawn[t] <- Nt[t,5] * S[5]      # Update spawner abundance
    SI[t] <- Ct[t,5] / E            # Update SI
  }
  
  Phi <- Spawn[nT] / Nt[nT,1]  # Ratio of spawners to immature at the final time step
  
  out <- list(Spawn = Spawn, SI = SI, Ct = Ct, Nt = Nt, Phi = Phi)
  return(out)
}

#this is our base model with no mr or fishing effort, its takes the output of the PrawnModel function and the spawners per immature prawn for each time step
Phi0 <- PrawnModel(MR = 0, E = 0)$Phi
#now we use seq to create a sequence of mr from 0 to 1 in increments of 0.01 to model the effect of mr
MR_values <- seq(0, 1, by = 0.01)
#empty vectors to store the spawning potential ratio and spawner index values for each mr
SPR_values <- numeric(length(MR_values))
SI_values <- numeric(length(MR_values))

#iterate through the mr values and calculate the spawning potential ratio and spawner index for each mr
for (i in seq_along(MR_values)) {
#set the mr value for the model
MR <- MR_values[i]
#initiate the model with the mr value and store the output
model_out <- PrawnModel(MR = MR, E = 400)
#store the spawners per immature prawn at the final time step for that mr
PhiF <- model_out$Phi
#store the spawning potential ratio for that mr, ratio of phiF to phi0
SPR_values[i] <- PhiF / Phi0
#store the spawner index for that mr, we use tail to get the last value in the vector
SI_values[i] <- tail(model_out$SI, 1)
}

print(SPR_values)
print(SI_values)
print(MR_values)

```